### 2019-03-13 Thesis/thesis.tex
Started writing thesis.
Some more progress.

@@

\section{Monads and Kleisli categories}

@@

The proof of Proposition \ref{pKleisli} essentially comes down to the following factorization result.  
If $f\from a \to b$ is a morphism in $\Kl_M$, then $f$ may be factorized as
\[
  f = a \xrightarrow{Jf}
  Mb \xrightarrow{\phi_b}
  b\,,
  \]
where we use `$f$' to refer both to the morphism $a\to b$ in $\Kl_M$ and to the underlying morphism $a \to Mb$ in $\C$.
Indeed, if we compute this composite inside $\C$, we get
\[
  a \xrightarrow{f}
  Mb \xrightarrow{u_{Mb}}
  MMb \xrightarrow{M\id}
  MMb \xrightarrow{m_b}
  Mb\,,
  \]
which is equal to $f$ by the coherence conditions on $m$ and $u$.
This means that the Kleisli category may be thought of as begin freely generated from the original category $\C$ and a multiplicative natural transformation $\phi$.

\begin{example}
  The morphisms in the Kleisli category for the nonempty powerset monad $\powerset_+$ on $\Set$ are functions $A \to \powerset_+B$, which can be thought of as nondeterministic functions.  
  Given a set $A$, the morphism $\phi_A\from \powerset_+A \to A$ in $\Kl_{\powerset_+}$ can be interpreted as a `nondeterministic choice' function that accepts a nonempty set of elements of $A$ and nondeterministically chooses one of them.
  The factorization then means that the category is freely generated over $\C$ by these nondeterministic choice morphisms.
\end{example}
\begin{example}
  Let $\C$ be a Cartesian closed category and let $z$ be some fixed object of $\C$.  
  Then the Kleisli category for the reader monad $R_z$ on $\C$ is generated over $\C$ by a natural transformation $\phi_y\from (z \to y) \to y$.  
  By the enriched Yoneda lemma, such a natural transformation is always given by precomposition with some fixed morphism $\ask\from 1 \to z$.  
  This means that $\Kl_{R_z}$ is suitable for modelling any situation in which we are generally working in $\C$, but need the ability to request a value of type $z$ (for example, a config file, a piece of user input or something else that isn't being passed into the function in question).
  \label{ExReaderMonadKleisli}
\end{example}

@@

\subsection{Extended example: game semantics of time complexity}
This is an extended example of using a reader monad, in the style of Example \ref{ExReaderMonadKleisli}.
Let $\G$ be the category of games and visible single-threaded strategies as in \cite{SamsonGuyIAActive}.  
So $\G$ is a sound and adequate model of Idealized Algol satisfying compact definability.
Let $\bC$ be the denotation of the command type $\com$ of Idealized Algol.  
Then the Kleisli category for the reader monad over $\bC$ is freely generated over $\G$ by a single morphism $1 \to \bC$.

Let us extend Idealized Algol with a single term, $\sleep$, of type $\com$.
This should be thought of as a command to the computer to pause for some fixed interval of time -- let us say a second -- before resuming execution.
By Theorem \ref{FunctionalCompletenessCcc}, our Kleisli category is Cartesian closed.
So it gives us a denotational semantics of IA+$\sleep$, in which the terms of IA are interpreted within $\G$ as in \cite{SamsonGuyIAActive} and the new term $\sleep$ is interpreted by the new morphism $1 \to \bC$.

Let us now give an operational semantics to terms of IA+$\sleep$.  
Let $\Gamma$ be an IA-context.
Then a \emph{store} $s$ over $\Gamma$ is a list of pairs $(v\mapsto x)$, where each $v$ is the name for a $\Var$-type variable occurring in $\Gamma$ and $x$ is a value held in $v$.  
We write $()$ for the empty store, $(s\vert v\mapsto x)$ for the store obtained from $s$ by adding or updating the variable $v$ to hold the value $x$ and $(s\vert v\mapsto \bot)$ for the store obtained by removing the variable $v$ from $s$.
Given disjoint stores $s,t$ (i.e., stores whose sets of variables are disjoint), we write $s+t$ for the store obtained by concatenating together $s$ and $t$.

Given a term $\Gamma\ts M$ in context, stores $s$,$s'$ over $\Gamma$, a value $c$ and a natural number $n$, we define a big-step relation
\[
  \Gamma,s\ts M\converges_n c,s'\,,
  \]
read as `in the context $\Gamma$, $s,M$ converges to $c,s'$ in time $n$'.  
The rule for the constant $\sleep$ is given by
\[
  \inferrule*
  { }
  {\Gamma,s\ts \sleep\converges_1 \skipp,s}\,,
  \]
indicating that execution of the $\sleep$ command takes up an additional time step.  
The other rules are the same as those given in \cite{SamsonGuyIAActive} for the $\converges$ relation, except that we have to modify them in order to take account of the new information we have added to the reduction relation (i.e., the time taken to converge).
For example:
\begin{mathpar}
  \inferrule*
  { }
  {\Gamma,s\ts c \converges_0 c,s}
  \and
  \inferrule*
  {\Gamma,s\ts M\converges_m\skipp,s'\\
  \Gamma,s'\ts N\converges_n\skipp,s''}
  {\Gamma,s\ts M;N\converges_{m+n}\skipp,s''}\,.
\end{mathpar}
If $P$ is a program (i.e., a term of type $\com$) then we write $P\converges_n$ as a shorthand for $,()\ts P \converges_n \skipp,()$.
We say that $P$ \emph{diverges} and write $P\diverges$ if there is no $n$ such that $P\converges_n$.

We now want a way to capture this operational behaviour within the denotational semantics.
First, we note the Computational Adequacy result proved for the game semantics of Idealized Algol.

\begin{proposition}[{\cite[21]{SamsonGuyIAActive}}]
  Let $P$ be a term of type $\com$.  
  Then $P\converges$ if and only if $\deno{P}\ne\bot$.
  \label{SamsonGuyAdequacy}
\end{proposition}

To make sense of this strategy, we need to know that $\G$ is enriched in algebraic cpos.  
Then $\bot$ is the bottom element of $\G(1,\bC)$.

Before proceeding further, we prove an important lemma.  
This lemma tells us that the operational semantics of a term of IA+$\sleep$ may be modelled within Idealized Algol; i.e., that IA+$\sleep$ is a conservative extension of the original language.

\begin{lemma}
  Let $P\from T$ be a term of IA+$\sleep$.
  Suppose that
  \[
    \Gamma,s\ts P\converges_n c,s'
    \]
  in IA+$\sleep$.

  Then for all $k\ge 0$, all contexts $\Delta$ disjoint from $\Gamma$, all stores $t$ over $\Gamma,\Delta$ disjoint from $s$ and all variables $v$ not in $\Gamma,\Delta$:

@@

    \Delta,\Gamma,v,(t{+}s\vert v \mapsto k)\ts P[v \gets \suc !v/\sleep] \converges c,(t{+}s\vert v \mapsto k+n)

@@

  in Idealized Algol.
  \label{IaSleepsSoundnessLemma}
\end{lemma}
\begin{proof}
  Induction on the derivation that $\Gamma,s\ts P\converges_n c,s'$.
  There are several possibilities for the last step in the derivation.
  \begin{description}
    \item[Structural congruence]
      Suppose that the last step in the derivation is

@@

        \inferrule*{P\equiv P'\\
        \Gamma,s\ts P' \converges_n \skipp, s'}
        {\Gamma,s\ts P\converges_n \skipp, s'}\,.

@@

      It is clear then by the definition of structural congruence that $P'[v\gets\suc !v/\sleep]\equiv P[v\gets\suc !v/\sleep]$.  
      So the claim follows by the inductive hypothesis applied to $P'$ and the structural congruence rule for Idealized Algol.
    \item[$\sleep$]
      Suppose that the last step in the derivation is

@@

        \inferrule*{ }
        {\Gamma,s\ts \sleep \converges_1 \skipp,s}\,.

@@

      Fix a context $\Delta$ disjoint from $\Gamma$ and a store $t$ over $\Gamma,\Delta$ that is disjoint from $s$.  
      Fix $k\ge n$.  
      Then it suffices to show that the following judgement may be proved within the operational semantics of Idealized Algol.
        \Delta,\Gamma,v,(t{+}s\vert v\mapsto k)\ts v \gets \suc !v \converges \skipp, (t{+}s\vert v\mapsto k+1)
      A derivation for this obvious fact is given in Figure \ref{SleepIADerivation}.

@@

                    \Delta,\Gamma,v,(t{+}s\vert v\mapsto k)\ts v \converges v,(t{+}s\vert v\mapsto k)
                  \Delta,\Gamma,v,(t{+}s\vert v\mapsto k) \ts !v \converges k,(t{+}s\vert v\mapsto k)
                \Delta,\Gamma,v,(t{+}s\vert v\mapsto k)\ts \suc !v \converges k+1,(t{+}s\vert v\mapsto k)
              } \\

@@

                \Delta,\Gamma,v,(t{+}s\vert v\mapsto k)\ts v \converges v,(t{+}s\vert v\mapsto k)

@@

              \Delta,\Gamma,v,(t{+}s\vert v\mapsto k)\ts v\gets\suc !v\converges\skipp,(t{+}s\vert v\mapsto k+1)

@@

          \caption{The crucial part of Lemma \ref{IaSleepsSoundnessLemma} pertaining to the $\sleep$ constant is a routine Idealized Algol derivation.}
          \label{SleepIADerivation}

@@

        \par\vspace{48pt}
            \inferrule*{\Delta,\Gamma_1,v,(t{+}s\vert v\mapsto k) \ts M_1[v \gets \suc !v/\sleep] \converges c_1,(t{+}s\vert v \mapsto k + n_1) \\
            \cdots \\
            \Delta,\Gamma_p,v,(t{+}s\vert v\mapsto k + n_1 + \cdots + n_{p-1}) \ts M_p[v \gets \suc !v/\sleep] \converges c_1,(t{+}s\vert v \mapsto k+n_1+\cdots+n_p)}
            {\Delta,\Gamma,v,(t{+}s^{(0)}\vert v \mapsto k + n_1 + \cdots + n_p)\ts M[v\gets \suc !v/\sleep] \converges c,(t{+}s^{(p)}\vert v \mapsto k)}
          \caption{The big-step rules of IA+$\sleep$ give rise to big-step rules of Idealized Algol via our translation.}
          \label{GeneralIADerivation}
        \caption{The proof of Lemma \ref{IaSleepsSoundnessLemma} works by translating big-step derivations from IA+$\sleep$ into big-step derivations of Idealized Algol.}
        \label{IADerivations}

@@

    \item[Canonical forms]
      Suppose that the last step in the derivation is
        \inferrule*{ }
        {\Gamma,s\ts_0 c\converges c,s}\,.
      Then, since $\sleep$ is not a canonical form, we have $c[v\gets \suc !v/\sleep]=c$.  
      Then we have the derivation
        \inferrule*{ }
        {\Delta,\Gamma,(t{+}s\vert v\mapsto k)\ts c \converges c,(t{+}s\vert v\mapsto k)}\,.
    \item[Sequencing]
      Suppose instead that the last step in the derivation is
        \inferrule*{\Gamma,s\ts M \converges_m \skipp, s'\\
        \Gamma,s'\ts N \converges_n \skipp, s''}
        {\Gamma,s\ts M;N \converges_{m+n} \skipp, s''}\,.
      Fix $k\ge 0$.  
      By induction on $M,N$, we have
      \begin{mathpar}
        \Delta,\Gamma,v,(t{+}s\vert v\mapsto k)\ts M[v\gets\pred !v/\sleep]\converges \skipp,(t{+}s'\vert  v\mapsto k+m)\,;
        \and
        \Delta,\Gamma,v,(t{+}s'\vert  v\mapsto k+m) \hspace{144pt} \par\vspace{-8pt} \\ \par \hspace{144pt} \ts N[v\gets\pred !v/\sleep]\converges \skipp,(t{+}s''\vert v\mapsto k+m+n)\,.
      \end{mathpar}
      It follows that
      \begin{mathpar}
        \Delta,\Gamma,v,(t{+}s\vert v\mapsto k) \hspace{180pt} \par\vspace{-8pt} \\ \par \hspace{180pt} \ts (M;N)[v\gets\pred !v/\sleep] \converges \skipp, (t{+}s''\vert v\mapsto k+m+n)\,,
      \end{mathpar}
      by the sequencing rule of Idealized Algol.
    \item[Remaining rules]
      Having given the rule for canonical forms and sequencing explicitly, we now give a general technique by which we can deal with all the remaining rules.

      Each rule is of the form
      \[
        \inferrule*{ \Gamma_1,s^{(0)}\ts z_1\converges_{n_1}\gamma_1,s^{(1)} \\
        \cdots \\
        \Gamma_p,s^{(p - 1)}\ts z_p \converges_{n_p} \gamma_p,s^{(p)}}
        { \Gamma,s^{(0)}\ts z \converges_{n_1+\cdots+n_p} \gamma,s^{(p)}}
        \]
      (for $p\in\{0,1,2,3\}$), where
      \[
        \inferrule*{ \Gamma_1,s^{(0)}\ts z_1\converges \gamma_p,s^{(1)} \\
        \cdots \\
        \Gamma_p,s^{(p - 1)}\ts z_p \converges \gamma_1,s^{(p)}}
        { \Gamma,s^{(0)}\ts z \converges \gamma,s^{(p)}}
        \]
      is a rule for Idealized Algol.

>>>>

\usepackage{lua-visual-debug}

@@

\section{Monads and Kleisli categories}
\ref{SecMonads}

@@

The proof of Proposition \ref{pKleisli} essentially comes down to the following factorization result.  
If $f\from a \to b$ is a morphism in $\Kl_M$, then $f$ may be factorized as
\[
  f = a \xrightarrow{Jf}
  Mb \xrightarrow{\phi_b}
  b\,,
  \]
where we use `$f$' to refer both to the morphism $a\to b$ in $\Kl_M$ and to the underlying morphism $a \to Mb$ in $\C$.
Indeed, if we compute this composite inside $\C$, we get
\[
  a \xrightarrow{f}
  Mb \xrightarrow{u_{Mb}}
  MMb \xrightarrow{M\id}
  MMb \xrightarrow{m_b}
  Mb\,,
  \]
which is equal to $f$ by the coherence conditions on $m$ and $u$.
This means that the Kleisli category may be thought of as being freely generated from the original category $\C$ and a multiplicative natural transformation $\phi$.

\begin{example}
  The morphisms in the Kleisli category for the nonempty powerset monad $\powerset_+$ on $\Set$ are functions $A \to \powerset_+B$, which can be thought of as nondeterministic functional programs.  
  Given a set $A$, the morphism $\phi_A\from \powerset_+A \to A$ in $\Kl_{\powerset_+}$ can be interpreted as a `nondeterministic choice' function that accepts a nonempty set of elements of $A$ and nondeterministically chooses one of them.
  The factorization then means that the category is freely generated over $\C$ by these nondeterministic choice morphisms.
\end{example}
\begin{example}
  Let $\C$ be a Cartesian closed category and let $z$ be some fixed object of $\C$.  
  Then the Kleisli category for the reader monad $R_z$ on $\C$ is generated over $\C$ by a natural transformation $\phi_y\from (z \to y) \to y$.  
  By the enriched Yoneda lemma, such a natural transformation is always given by precomposition with some fixed morphism $\ask_z\from 1 \to z$.  
  This means that $\Kl_{R_z}$ is suitable for modelling any situation in which we are generally working in $\C$, but need the ability to request a value of type $z$ (for example, a config file, a piece of user input or something else that isn't being passed into the function in question).
  \label{ExReaderMonadKleisli}
\end{example}

@@

\subsection{Denotational Semantics}

From now till the end of the chapter, we fix an (order-enriched) Cartesian closed category $\G$ that admits a denotational semantics of Idealized Algol satisfying Computational Adequacy and in which every compact element is definable.
The prototypical example, of course, will be the category of games and visible strategies, but we will not exploit any properties of this model beyond the ones we have already mentioned, mentioning it only in examples where appropriate.

Let $X\in \{\bB,\bN,\bC\}$ be a set that has an interpretation as an Idealized Algol type $X$, and write $X$ also for the corresponding object of $\G$.
Write $\G_X$ as a shorthand for $\Kl_{R_X}\G$, the Kleisli category for the reader monad on $\G$ corresponding to the object $X$.  
The purpose of the rest of this chapter will be to define a new language, give it a denotational semantics in $\G_X$, and prove a full abstraction result for this denotational semantics.

\newcommand{\IAX}{{IA${}_X$}\xspace}
\begin{definition}[{The language \IAX}]
  The language \IAX is formed by taking Idealized Algol, and adding to it a new constant
  \[
    \ask_X
    \]
  with typing rule
  \[
    \inferrule{ }{\Gamma \ts \ask_X \from X}\,.
    \]
\end{definition}

From Proposition \ref{pKleisli}, we know that there is a distinguished natural transformation $\phi_A \from (X \to A) \to A$ in $\G_X$; in particular, we have a morphism
\[
  \phi = \phi_X(\id_X) \from 1 \to X\,,
  \]
which will be the denotation of the term $\choose_X$.
Together with the existing denotational semantics of Idealized Algol within $\G$, this gives us an inductively defined denotational semantics of \IAX within $\G_X$.

Clearly any term-in-context of \IAX is of the form
\[
  \Gamma \ts M[\ask_X/x]\from T\,,
  \]
where
\[
  \Gamma,x\from X \ts M \from T
  \]
is a judgement of Idealized Algol.
Given such a term-in-context, we know that the denotation of
\[
  \Gamma\ts (\lambda x.M) \ask_X \from T
  \]
is given by the composite
\[
  1 \xrightarrow{\phi}
  X \xrightarrow{\deno{\Gamma,x \ts M}}
  \deno{T}\,.
  \]
Now this last term is $\beta$-equivalent to our original term-in-context $\Gamma\ts M$.  
Since $\G_X$ is Cartesian closed (by Theorem \ref{FunctionalCompletenessCcc}), the $\beta$ rule is valid in $\G_X$, and this means that the composite above is an alternative definition of the denotation of $\Gamma\ts M$.

\subsection{Operational Semantics}

We now define the operational semantics of \IAX and prove a computational adequacy result for our denotational semantics.

\begin{definition}[Operational semantics of \IAX]
  Let $X^*$ be the free monoid on the set $X$; i.e., the set of all finite strings of elements of $X$.
  Given $u,v\in X^*$ we shall write $u\cat v$ for their product in $X^*$; i.e., the concatenation of the two strings.
  We shall write $\epsilon$ for the unit in $X^*$; i.e., the empty string.

  If $u\in X^*$,we write $|u|$ for the length of $u$.
  If $0\le n < |u|$, then we write $u^{(n)}$ for the corresponding element of $u$, numbering from $0$.

  We inductively define a relation $\Gamma,s\ts M\converges_u c,s'$, where $\Gamma$ is a $\Var$-context, $M,c$ are terms of \IAX with all free variables in $\Gamma$, where $c$ is an IA canonical form, $s,s'$ are $\Gamma$-stores and $u\in X^*$.
  The definition of this relation is shown in Figure \ref{FigIaxOpSem}.

  \begin{figure}
    \begin{mathpar}
      \inferrule*{ }{\Gamma,s\ts c \converges_\epsilon c,s}
      \and
      \inferrule*{\Gamma,s \ts M \converges_u \lambda x.M',s' \\ \Gamma,s' \ts M'[N/x] \converges_v c,s''}{\Gamma,s \ts MN \converges_{u\cat v} c,s''}
      \and
      \inferrule*{\Gamma,s \ts M(\Y M) \converges_u c,s'}{\Gamma,s \ts \Y M \converges_u c,s'}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u n,s'}{\Gamma,s\ts \suc M \converges_u n+1,s'}
      \\\and
      \inferrule*{\Gamma,s\ts M \converges_u n+1,s'}{\Gamma,s\ts \pred M \converges_u n,s'}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u 0,s'}{\Gamma,s\ts \pred M \converges_u 0,s'}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u \skipp,s' \\ \Gamma,s'\ts N \converges_v c,s''}{\Gamma,s \ts M;N \converges_{u\cat v} c,s''}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u \true,s' \\ \Gamma,s' \ts N \converges_v c,s''}{\Gamma,s \ts \If M \Then N \Else P \converges_{u\cat v} c,s''}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u \false,s' \\ \Gamma,s' \ts P \converges_v c,s''}{\Gamma,s \ts \If M \Then N \Else P \converges_{u\cat v} c,s''}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u 0,s' \\ \Gamma,s' \ts N \converges_v c,s''}{\Gamma,s \ts \IfO M \Then N \Else P \converges_{u\cat v} c,s''}
      \and
      \inferrule*{\Gamma,s\ts M \converges_u n+1,s' \\ \Gamma,s' \ts P \converges_v c,s''}{\Gamma,s \ts \IfO M \Then N \Else P \converges_{u\cat v} c,s''}
      \and
      \inferrule*[right=$x\in\Gamma$]{\Gamma,s\ts E \converges_u n,s' \\ \Gamma,s' \ts V \converges_v x,s''}{\Gamma,s\ts V \gets E \converges_{u\cat v} \skipp,(s''\vert x \mapsto n)}
      \and
      \inferrule*[right={$s'(x)=n$}]{\Gamma,s\ts V \converges_u x,s'}{\Gamma,s\ts !V \converges_u n,s'}
      \and
      \inferrule*{\Gamma,x\from\Var,(s\vert x\mapsto 0)\ts M \converges_u c,(s'\vert x\mapsto n)}{\Gamma,s\ts \neww \lambda x.M \converges_u c,s'}
      \and
      \inferrule*{\Gamma,s\ts E \converges_u n,s' \\ \Gamma,s'\ts V \converges_v \mkvar W R,s'' \\ \Gamma,s'' \ts Wn \converges_w \skipp,s'''}
      {\Gamma,s \ts V\gets E \converges_{u\cat v\cat w} \skipp,s'''}
      \and
      \inferrule*{\Gamma,s\ts V \converges_u \mkvar W R,s' \\ \Gamma,s'\ts R \converges_v n,s''}{\Gamma,s\ts !V \converges_{u\cat v} n,s''}
      \\\and
      \inferrule*[right=$x\in X$]{ }{\Gamma,s\ts \ask_X \converges_x x,s}
    \end{mathpar}
    \caption[Operational semantics for \IAX]{Operational semantics for \IAX.  
    All the rules except the last one are deterministic and may be obtained from the corresponding rules of Idealized Algol by suitably annotating the $\converges$ relation with sequences from $X^*$.}
    \label{FigIaxOpSem}
  \end{figure}

  We can define this semantics in an alternative, indirect way.  
  Note that each rule from ordinary Idealized Algol takes the form
  \[
    \inferrule{\Gamma,s^{(0)}\ts M_1\converges c_1,s^{(1)} \\ \cdots \\ \Gamma,s^{(n-1)}\ts M_n\converges c_n,s^{(n)}}
    {\Gamma,s^{(0)} \ts M \converges c,s^{(n)}}\,,
    \]
  Here, we have interpreted each IA rule as an infinite scheme of rules ranging over the different terms $M_i,M$ that the rule can apply to.
  We first extend this rule to a rule for \IAX, by allowing the $M_i,M$ to range over terms of \IAX.
  We then replace the rule with the new rule
  \[
    \inferrule{\Gamma,s^{(0)}\ts M_1 \converges_{u_1} c_1,s^{(1)} \\ \cdots \\ \Gamma,s^{(n-1)}\ts M_n \converges_{u_n} c_n,s^{(n)}}
    {\Gamma,s^{(0)} \ts M \converges_{u_1 \cat \cdots \cat u_n} c,s^{(n)}}\,,
    \]
  to give us an operational rule for \IAX (if $n=0$, then we treat the empty string $\epsilon$ as the empty concatenation).
  Lastly, we add the rule for the new constant $\ask_X$:
  \[
    \inferrule*[right=$x\in X$]{ }{\Gamma,s \ts \ask_X \converges_x x,s}\,.
    \]
  This rule is the only nondeterministic one in our language, as well as being the only one in which the sequence annotating the $\converges$ symbol at the bottom is not formed from concatenating together the sequences on the top.
\end{definition}

\begin{example}
  If $X=\bC$, then, since $X$ has a single element, a sequence $n$ of elements of $X$ may be identified with its length $n$.
  In this case, the language \IAX gives us a way to model time complexity, and the term $\ask_X$ may be considered as a constant $\sleep\from \com$ whose semantics is to wait for some fixed period of time before continuing.  
  In this case, 
  \[
    \Gamma,s\ts M \converges_n c,s'
    \]
  is interpreted to say that `$M$ converges to $c$ in time $n$'.
\end{example}
\begin{example}
  If $X\in\{\bB,\bN\}$, then the language \IAX gives us a way to model nondeterminism, where $\ask_X$ behaves as a \emph{nondeterministic oracle}; i.e., a device that nondeterministically returns an element of $X$.

  If $X=\bB$ then we have a model of binary (i.e., finite) nondeterminism, whereas if $X=\bN$ then we have a model of countable nondeterminism.

  We interpret the relation
  \[
    \Gamma,s\ts M \converges_u c,s'
    \]
  as saying that $M$ converges to $c$ in the case that the sequence of values returned by the nondeterministic oracle is given by the sequence $u$.
\end{example}

\subsection{Soundness}
\label{SecSoundness}

To prove our adequacy result for the operational semantics of \IAX, we first give some definitions.

\begin{definition}
  Fix some constant value $\top \in X$ (the precise value does not matter).
  We inductively define terms in context $\tr_u\from \nat \to X$ of \emph{ordinary deterministic} Idealized Algol for each $u\in X^*$ as follows.
  \begin{mathpar}
    \tr_\epsilon = \lambda n.\top
    \and
    \tr_{xu} = \lambda n.\neww (\lambda v.v\gets n;\IfO !v \Then x \Else \tr_u (\pred !v))
  \end{mathpar}
\end{definition}

\begin{proposition}
  Let $u \in X^*$ and let $n<|u|$.  
  Then it is possible to deduce that
  \[
    \inferrule{\Gamma,s \ts M \converges n,s'}{\Gamma,s \ts \tr_u M \converges u^{(n)},s'}
    \]
  in Idealized Algol.
  \label{PropTr}
\end{proposition}
\begin{proof}
  Induction on $|u|$ and on $n$.
  Since $n<u$, $u$ must be non-empty, of the form $xu'$.

  Suppose $n=0$.  
  Then $u^{(n)}=x$, and we have a derivation of $\Gamma,s\ts \tr_{xu} M\converges x,s'$ from $\Gamma,s\ts M\converges n,s'$ as shown in Figure \ref{FigFirstTermOfSequence}.

  Now suppose that $n=m+1$.  
  Then $(xu)^{(m+1)}=u^{(m)}$.  
  Then we have a derivation of $\Gamma,s\ts \tr_{xu} M \converges u^{(m)},s'$ from $\Gamma,s\ts M\converges n,s'$ in Figure \ref{FigGeneralTermOOfSequence}, using the inductive hypothesis to tell us that we may derive

@@

    \inferrule*{\Gamma,v,(s'\vert v\mapsto m+1) \ts \pred !v \converges m,(s'\vert v \mapsto m+1)}
    {\Gamma,v,(s'\vert v \mapsto m+1) \ts \tr_u(\pred !v) \mapsto u^{(m)},(s'\vert v\mapsto m+1)}\,.\qedhere

@@

    \tiny

@@

                \Gamma,v,(s\vert v \mapsto 0) \ts M \converges 0,(s'\vert v\mapsto 0)
                \\
                  \Gamma,v,(s'\vert v \mapsto 0)\ts v \converges v,(s'\vert v\mapsto 0)
                \Gamma,v,(s\vert v\mapsto 0) \ts v \gets M \converges \skipp,(s'\vert v\mapsto 0)
              \\
                    \Gamma,v,(s'\vert v\mapsto 0) \ts v \converges v,(s'\vert v\mapsto 0)
                  \Gamma,v,(s'\vert v\mapsto 0) \ts !v \converges 0,(s'\vert v\mapsto 0)
                \\
                  \Gamma,v,(s'\vert v\mapsto 0) \ts x \converges x,(s'\vert v\mapsto 0)
                \Gamma,v,(s'\vert v\mapsto 0) \ts \IfO !v \Then x \Else \tr_u(\pred !v) \converges x,(s'\vert v\mapsto 0)
              \Gamma,v,(s\vert v\mapsto 0) \ts v\gets M;\IfO !v \Then x \Else \tr_u(\pred !v) \converges x,(s'\vert v\mapsto 0)
            \Gamma,s\ts \neww (\lambda v.v\gets M;\IfO !v \Then x \Else \tr_u(\pred !v)) \converges x,s'
          \Gamma,s\ts \lambda n.\neww (\lambda v.v\gets n;\IfO !v \Then x \Else \tr_u(\pred !v)) M \converges x,s'

@@

      \caption{IA derivation that if $M\converges 0$ then $\tr_u M$ converges to the first element of the sequence $u$.}
      \label{FigFirstTermOfSequence}
    \par\vspace{24pt}

@@

                \inferrule*[rightskip=30pt]
                  \Gamma,v,(s'\vert v\mapsto 0) \ts v \converges v,(s'\vert v\mapsto 0)
                \\\\
                \inferrule*[leftskip=15pt]{}{\Gamma,v,(s\vert v\mapsto 0) \ts M \converges m+1, (s'\vert v\mapsto 0)}
                \Gamma,v,(s\vert v\mapsto 0) \ts v \gets M \converges \skipp,(s'\vert v\mapsto m+1)
              \\
                    \Gamma,v,(s'\vert v\mapsto m+1) \ts v \converges v,(s'\vert v\mapsto m+1)
                  \Gamma,v,(s'\vert v\mapsto m+1) \ts !v \converges m+1,(s'\vert v\mapsto m+1)
                \\
                        \Gamma,v,(s'\vert v\mapsto m+1) \ts v \converges v,(s'\vert v\mapsto m+1)
                      \Gamma,v,(s'\vert v\mapsto m+1) \ts !v \converges m+1,(s'\vert v\mapsto m+1)
                    \Gamma,v,(s'\vert v\mapsto m+1) \ts \pred !v \converges m,(s'\vert v\mapsto m+1)
                  \Gamma,v,(s'\vert v\mapsto m+1) \ts \tr_u(\pred !v) \converges u^{(m)}, (s'\vert v \mapsto m+1)
                \Gamma,v,(s'\vert v\mapsto m+1) \ts \IfO !v \Then x \Else \tr_u(\pred !v) \converges u^{(m)}, (s'\vert v\mapsto m+1)
              \Gamma,v,(s\vert v\mapsto 0) \ts v \gets M;\IfO !v \Then x \Else \tr_u(\pred !v) \converges u^{(m)}, (s'\vert v \mapsto m+1)
            \Gamma,s \ts \neww (\lambda v.v\gets M;\IfO !v \Then x \Else \tr_u(\pred !v)) \converges u^{(m)},s'
          \Gamma,s\ts \lambda n.\new (\lambda v.v\gets n;\IfO !v \Then x \Else \tr_u(\pred !v)) M \converges u^{(m)},s'

@@

        \caption{IA derivation that if $M\converges m+1$ then $\tr_{u}M$ converges to the $m+1$-th element of the sequence $u$.}
        \label{FigGeneralTermOOfSequence}
    \par\vspace{24pt}

@@

@@

                  \Gamma,v,(s\vert v\mapsto k) \ts v \converges v,(s\vert v\mapsto k)

@@

                \Gamma,v,(s\vert v\mapsto k) \ts !v \converges k,(s\vert v\mapsto k)
              \Gamma,v,(s\vert v\mapsto k) \ts \suc !v \converges k+1,(s\vert v\mapsto k)
            \\
              \Gamma,v,(s\vert v\mapsto k) \ts v\converges v,(s\vert v\mapsto k)
            \Gamma,v,(s\vert v\mapsto k) \ts v\gets \suc !v \converges \skipp,(s\vert v\mapsto k+1)
          \\
            \inferrule*[right={Prop. \ref{PropTr}}]
                  \Gamma,v,(s\vert v\mapsto k+1) \ts v\converges v,(s\vert v\mapsto k+1)
                \Gamma,v,(s\vert v\mapsto k+1) \ts !v \converges k+1,(s\vert v\mapsto k+1)
              \Gamma,v,(s\vert v\mapsto k+1) \ts \tr_w !v \converges x,(s\vert v\mapsto k+1)
          \Gamma,v,(s\vert v\mapsto k) \ts v\gets \suc !v;\tr_w !v \converges x,(s\vert v\mapsto k+1)

@@

      \caption{IA derivation that $(x\mapsto k),v \gets \suc !v;\tr_w !v$ converges to the $k+1$-th term of $w$.}
      \label{FigTheOneForTheLemma}

@@

    \normalsize

@@

\end{proof}

We need a small lemma to help us deal with substitution.

\begin{lemma}
  Let
  \[
    \inferrule{\Gamma,s^{(0)}\ts M_1\converges c_1,s^{(1)} \\ \cdots \\ \Gamma,s^{(n-1)}\ts M_n\converges c_n,s^{(n)}}
    {\Gamma,s^{(0)} \ts M \converges c,s^{(n)}}
    \]
  be an inference, where the $M_i$,$M$ are terms of \IAX and the whole inference satisfies one of the patterns of the Idealized Algol rules.  
  Let $Q$ be a fixed term of type $X$.  
  Then
  \[
    \inferrule{\Gamma,s^{(0)}\ts M_1[Q/\ask_X]\converges c_1,s^{(1)} \\ \cdots \\ \Gamma,s^{(n-1)}\ts M_n[Q/\ask_X]\converges c_n,s^{(n)}}
    {\Gamma,s^{(0)} \ts M[Q/\ask_X] \converges c,s^{(n)}}\,,
    \]
  is a valid inference of Idealized Algol.
  \label{LemFirstSubstitution}
\end{lemma}
\begin{proof}
  The real reason this is true is that the term $\ask_X$ is not mentioned anywhere in the IA rules, so substitution of the term $N$ for $\ask$ could not possibly break the pattern.  
  Formally, we can show this by inspection on each of the different rules.  
  For instance, if the original rule is the one for sequencing:
  \[
    \inferrule{\Gamma,s\ts M \converges \skipp,s' \\ \Gamma,s'\ts N \converges c,s''}
    {\Gamma,s \ts M;N \converges c,s''}\,,
    \]
  then we have $(M;N)[Q/\ask_X] = M[P/\ask_X];N[P/\ask_X]$ and the inference
  \[
    \inferrule{\Gamma,s\ts M[Q/\ask_X] \converges \skipp,s' \\ \Gamma,s'\ts N[Q/\ask_X]\converges c,s''}
    {\Gamma,s \ts M[Q/\ask_X];N[P/\ask_X] \converges c,s''}
    \]
  is still a valid instance of the sequencing rule.
\end{proof}

We can now state and prove our soundness lemma.

\begin{lemma}
  Suppose that
  \[
    \Gamma,s\ts M\converges_u c,s'
    \]
  in \IAX.
  Fix $k\in\bN$ and let $w\in X^*$ be a sequence such that $u$ is a subsequence of $w$ starting at position $k+1$ (i.e., $u^{(j)}=w^{(k+j+1)}$ for each $j=0,\cdots,|u|-1$).
  Then
  \[
    \Gamma,v\from\Var,(s\vert v\mapsto k) \ts M[v\gets \suc!v;\tr_w !v/\ask_v] \converges c,(s'\vert v\mapsto k + |u|)
    \]
  in Idealized Algol.
  \label{LemSoundness}
\end{lemma}
\begin{proof}
  Structural induction on the derivation.
